Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-19 paper/paper.tex
stuff

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction for
structuring recursion while ensuring termination and providing reasoning about
program equivalences as well as deriving optimisations which has been
successfully applied to functional programming. Formalising recursion schemes in
a type theory offers additional termination guarantees, but it often requires
compromises affecting the resulting code, such as imposing performance
penalties, requiring the assumption of additional axioms, or introducing unsafe
casts into extracted code (e.g. \ocaml{Obj.magic} in OCaml).

@@

However, when used in the context of languages with general recursion like
(e.g.) Haskell, recursion schemes cannot ensure termination, but while
formalising recursion schemes in a type theory does provide stronger termination
guarantees, to the best of our knowledge, not many attempts have been made at
mechanising structured recursion schemes.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is problematic since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, most uses of extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation to our knowledge of
\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
extraction of idiomatic OCaml code. The full mechanisation can be found on Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}).

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
As an example of how can we use program calculation techniques in our 
framework, we show how another traversal can be fused into the
divide-and-conquer algorithm using the laws of hylomorphisms.
Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge tsplit.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting, and \coq{hylo_fusion_l} to fuse
\coq{times_two} into the RHS hylomorphism in \coq{qsort_times_two}.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is \coq{hylo (merge \o natural times_two) tsplit}. In this definition,
\coq{natural} defines a natural transformation by applying \coq{times_two} to
the shapes, and \coq{times_two} multiplies every pivot in the Quicksort tree by
two. Our formalisation contains a proof that \coq{natural} is indeed a natural
transformation, which relies on the fact that it preserves the structure of the
shapes and, therefore, the validity of the positions.  The extracted OCaml code
is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}

>>>>

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction for
structuring recursion while ensuring termination and providing reasoning about
program equivalences, as well as deriving optimisations which have been
successfully applied to functional programming. Formalising recursion schemes in
a type theory offers additional termination guarantees, but it often requires
compromises affecting the resulting code, such as imposing performance
penalties, requiring the assumption of additional axioms, or introducing unsafe
casts into extracted code (e.g. \ocaml{Obj.magic} in OCaml).

@@

However, when used in the context of languages with general recursion like
 Haskell, recursion schemes cannot ensure termination, but while
formalising recursion schemes in a type theory does provide stronger termination
guarantees, to the best of our knowledge, not many attempts have been made at
mechanising structured recursion schemes.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is problematic since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, in Abreu et al.'s approach, extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first Coq formalisation to our knowledge of
\emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
extraction of idiomatic OCaml code. The full mechanisation can be found on Github\footnote{https://github.com/dcastrop/coq-hylomorphisms}. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code (see \ocaml{qsort} in Section~\ref{sec:examples}).

@@

% Putting everything together, Mergesort is defined as the following
% hylomorphism:
% \begin{coqcode}
% Definition msort := hylo merge tsplit.
% \end{coqcode}
% Note, however, that we could have composed the catamorphism and anamorphism,
% and used the deforestation result to optimise the intermediate binary tree
% away:
% \begin{coqcode}
% Definition msort : Ext (cata merge \o rana tsplit).
% calculate. rewrite (deforest l_in_out). simpl. reflexivity. Defined.
% \end{coqcode}
% In both cases, the result is the same, and it extracts to the usual divide and
% conquer definition of Mergesort in OCaml:
% \begin{ocamlcode}
% let rec msort x = match x with
% | [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
%      let (l0, r) = splitL x [] [] in
%      let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
%      mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
% \end{ocamlcode}
\subsubsection{Fusing a divide-and-conquer computation}
As an example of how can we use program calculation techniques in our 
framework, we show how another traversal can be fused into the
divide-and-conquer algorithm using the laws of hylomorphisms.
Suppose that we map a function to the result of sorting the list. We can
use our framework to fuse both computations. In particular, consider
the following definition:
\begin{coqcode}
Definition qsort_times_two := Lmap times_two \o hylo merge c_split.
\end{coqcode}
Here, \coq{Lmap times_two} is a list map function defined as a hylomorphism,
and \coq{times_two} multiplies every element of the list by two.
We can use Coq's generalised rewriting, and \coq{hylo_fusion_l} to fuse
\coq{times_two} into the RHS hylomorphism in \coq{qsort_times_two}.
After applying hylo fusion and the necessary rewrites, the hylomorphism that we
extract is \coq{hylo (merge \o natural times_two) c_split}. In this definition,
\coq{natural} defines a natural transformation by applying \coq{times_two} to
the shapes, and \coq{times_two} multiplies every pivot in the Quicksort tree by
two. Our formalisation contains a proof that \coq{natural} is indeed a natural
transformation, which relies on the fact that it preserves the structure of the
shapes and, therefore, the validity of the positions.  The extracted OCaml code
is a single recursive traversal:
\begin{ocamlcode}
let rec qsort_times_two = function | [] -> []
| h :: t -> let (l, r) = partition (fun x0 -> leb x0 h) t in
            let x0 = fun p -> qsort_times_two (match p with
                                 | Lbranch -> l | Rbranch -> r) in
            app (x0 Lbranch) ((mul (Uint63.of_int (2)) h) :: (x0 Rbranch))
\end{ocamlcode}
